Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-01-15 joss/paper.Rmd
Edits joss manuscript

The growth of open-source statistical software programming has been
explosive in the last decade. In particular, the statistical
programming language ``R`` has grown exponentially to become one of
the top ten programming languages in use today. Recently, concerns
have arisen over the reproducibility of scientific research
[@Peng2011 @Baker2016 @Stodden2018] and the potential issues stemming
from the complexity and fragility of analytical software
[@Pasquier2017 @Chen2018]. There is now a recognition that simply
making the code open is not enough, and that there is a need for
improvements to documentation and transparency [@Chen2018]. From this
perspective, tools that can lower the time and energy required to
re-factor and streamline analytical scripts could have a significant
impact on scientific reproducibility across all disciplines
[@Visser2015]. Supporting this objective, we have created ``Rclean``
which automatically reduces a script to the parts that are
specifically relevant to a research product (e.g.  a blog, academic
talk or research article).

At it's root R is a statistical programming language. That is, it was
designed for use in analytical workflows. As such, the majority of the
R community is focused on producing code for idiosyncratic projects
that are results oriented. Also, R's design is intentionally at a
level that abstracts many aspects of programming that would otherwise
act as a barrier to entry for many users. This is good in that there
are many people who use R to their benefit with little to no formal
training in computer science or software engineering. However, these
same users are also frequently frustrated by code that is fragile,
buggy and complicated enough to quickly become obtuse even to
themselves in a very short amount of time. In addition, when scripts
take an extremely long time to execute, being able to reduce
unnecessary analyses can help increase computation efficiency.

More often then not, when someone is writing an R script, they are
intent on getting a set of results. This set of results is always a
subset of a much larger set of possible ways to explore a dataset, as
there are many statistical approaches and tests, let alone ways to
create visualizations and other representations of patterns in
data. This commonly leads to lengthy, complicated scripts from which
researchers manually subset results, but never refactor (i.e. reduce
to the final subset). In part, this is enabled by a lack of a proper
version control system, and in order to record their process and not
lose work, the entire process remains in a single or several
scripts. Although *Rclean* is not designed to fix the latter, it can
help with the former issue, once an appropriate versioning system is
adopted (e.g. git or subversion).

@@

This process relies on the
generation of data provenance [@Carata2014], which is a formal
representation of the execution of a computational process
(https://www.w3.org/TR/prov-dm/), to rigorously determine the the
unique computational pathway from inputs to results. However, as the
intended user is a researcher conducting analyses, the process is
abstracted and only the minimum information is required and presented
to the user to streamline the process of creating "cleaner" code. The
output generated by ``RClean`` is the minimum and sufficient code
needed to generate the chosen result.

The workhorse behind *Rclean* is data provenance. Here, when we refer
to provenance we are talking about a formalized representation of the
computational process that produced some data. Data is used in a broad
sense, not just data that were collected in a research project. There
are multiple approaches to collecting data provenance, but *Rclean*
uses "prospective" provenance, which analyzes code and uses language
specific information to predict the relationship among processes and
data objects. *Rclean* relies on a library called *CodeDepends* to
gather the prospective provenance for each script. For more
information on the mechanics of the *CodeDepends* package, see
[@Lang2019]. To get an idea of what data provenance is, take a look at
the `code_graph` function. The plot that it generates is a graphical
representation of the prospective provenance generated for *Rclean*.

```{R prov-graph}

code_graph(script)

## The API

The package's main function is `clean`.

code_graph
get_libs
get_vars
keep

>>>>

The growth of programming in the sciences has been explosive in the
last decade. In particular, the statistical programming language ``R``
has grown exponentially to become one of the top ten programming
languages in use today. Recently, concerns have arisen over the
reproducibility of scientific research [@Peng2011 @Baker2016
@Stodden2018] and the potential issues stemming from the complexity
and fragility of analytical software [@Pasquier2017 @Chen2018]. There
is now a recognition that simply making the code open is not enough,
and that there is a need for improvements to documentation and
transparency [@Chen2018].

At it's root R is a statistical programming language, that is, it was
designed for use in analytical workflows. As such, the majority of the
R community is focused on producing code for idiosyncratic projects
that are results oriented. Also, R's design is intentionally at a
level that abstracts many aspects of programming that would otherwise
act as a barrier to entry for many users. This is good in that there
are many people who use R to their benefit with little to no formal
training in computer science or software engineering. However, these
same users are also frequently frustrated by code that is fragile,
buggy and complicated enough to quickly become obtuse (even to
themselves) in a very short amount of time, which leads to frequently
re-writing code unecessarily. Also, when scripts take an extremely
long time to execute, reducing unnecessary analyses can increase
computational efficiency.

From this perspective, tools that can lower the time and energy
required to re-factor and streamline analytical scripts, and otherwise
help to "clean" code will have a significant impact on scientific
reproducibility across all disciplines [@Visser2015]. To support this
objective, we have created ``Rclean`` which automatically reduces a
script to the parts that are specifically relevant to a research
product (e.g.  a blog, academic talk or research article). Here, we
detail the structure of the package's API, describe the general
workflow illustrated by an example use case and provide some
background on how the underlying use of data provenance enables the
package.

@@

The main goal of *Rclean* is to provide a means to simplify code. To
keep the process simple and straight-forward, the API has been kept to
a minimum set of functions, which enable a user to conduct the basic
workflow of getting information about the posible results in a script,
obtaining the minimum code for a set of results and then creating a
new "cleaned" script or other software (e.g. a function, reproducible
example, web-app, etc.).

The package's main functions are `clean` and `keep`. When provided a
file path to a script and the name of a result (or a set of results),
`clean` analyzes the script's code and extracts the lines of code
required to produce the results. This code can then be passed to the
`keep` function, which can either write the code to disk or copy the
code to the user's clipboard (if no ouptput file path is supplied) and
the user can paste the code into another location (e.g. a script
editor).

In the process of cleaning a script, it is likely that a user might
require some help analyzing the script. There are several functions to
help. The `get_vars` function will return a list of possible results
for a given script at a supplied file path. This is obviously an
important step, and justsifiably, the default behavior of the `clean`
function is to run `get_vars` if no results are supplied. To help with
limiting and checking the selection of results, the `code_graph` function
creates a network graph of the relationships among the various
results and lines of code in the script. Last, the `get_libs` function
can be used to detect the packages that a given script depends on,
which it will return as coded library calls that can be inserted into
a cleaned script.
